---
title: "Smoking and Birth Weight Analysis"
format: html
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(janitor)
library(broom)
library(gtsummary)
library(here)

theme_set(theme_minimal())

age_min <- 18
bad_ids <- 1:3
bw_cutoff <- 2500
preterm_cutoff <- 37
```

## Overview

This report analyzes the relationship between smoking during pregnancy and birth outcomes after restricting to mothers aged `r age_min` years and older and excluding the records flagged during data review.

```{r}
#| label: read-clean-data
births <- read_csv(
  here("data", "births.csv"),
  show_col_types = FALSE
) |>
  clean_names() |>
  filter(
    mat_age >= age_min,
    !id %in% bad_ids
  ) |>
  mutate(
    low_bw = if_else(birth_weight < bw_cutoff, 1, 0),
    preterm = if_else(ges_age < preterm_cutoff, 1, 0),
    smoke = if_else(smoker == "Yes", 1, 0)
  ) |>
  filter(!is.na(birth_weight))
```

The analysis dataset includes `r nrow(births)` births. There were `r sum(births$low_bw)` low-birth-weight births and `r sum(births$preterm)` preterm births.

## Table 1

```{r}
#| label: tbl-table-1
#| tbl-cap: "Participant characteristics by smoking status"
table1 <- births |>
  select(smoker, mat_age, mat_bmi, parity, education, low_bw, preterm) |>
  tbl_summary(by = smoker) |>
  add_overall()

table1
```

The median maternal age was `r inline_text(table1, variable = "mat_age", column = "stat_0")` overall.

## Birth Weight

```{r}
#| label: fig-birthweight
#| fig-cap: "Birth weight by smoking status"
ggplot(births, aes(x = smoker, y = birth_weight, fill = smoker)) +
  geom_boxplot() +
  labs(
    x = "Smoker",
    y = "Birth weight (g)"
  ) +
  theme(legend.position = "none")
```

## Models

```{r}
#| label: fit-models
low_bw_model <- glm(
  low_bw ~ smoke + mat_age + mat_bmi + as.factor(education) + parity,
  data = births,
  family = binomial
)

preterm_model <- glm(
  preterm ~ smoke + mat_age + mat_bmi + as.factor(education) + parity,
  data = births,
  family = binomial
)
```

```{r}
#| label: tbl-low-bw-model
#| tbl-cap: "Odds ratios for low birth weight"
tbl_regression(low_bw_model, exponentiate = TRUE)
```

```{r}
#| label: tbl-preterm-model
#| tbl-cap: "Odds ratios for preterm birth"
tbl_regression(preterm_model, exponentiate = TRUE)
```

The adjusted odds ratio for low birth weight comparing smokers with non-smokers was `r inline_text(tbl_regression(low_bw_model, exponentiate = TRUE), variable = "smoke")`.
